---
title: "Draft: Data Documentation"
author: Max Buettenback
format: html
bibliography: references.bib
---

# Data Documentation {.unnumbered}

## Goal {.unnumbered}

The goal of this report is to estimate the risk of a bridge collapse in Omaha, Nebraska over the next five years.

Risk = Probability of failure × Consequence of failure

---

# Source 1: National Bridge Inventory {.unnumbered}

The data comes from the Federal Highway Administration's National Bridge Inventory [@fhwa_nbi]. The main variables used from this dataset are shown in @tbl-nbi.

## Important Variables {.unnumbered}

```{r}
#| echo: false
#| label: tbl-nbi
#| tbl-cap: "Variables used from the National Bridge Inventory dataset."

library(knitr)

kable(
  data.frame(
    Variable = c("Year Built", "Deck Rating", "Superstructure Rating",
                 "Substructure Rating", "Sufficiency Rating", "ADT"),
    Data_Type = c("Numeric", "Ordinal (0–9)", "Ordinal (0–9)",
                  "Ordinal (0–9)", "Numeric (0–100)", "Numeric"),
    Description = c("Used to calculate bridge age",
                    "Condition of the bridge deck",
                    "Condition of load-bearing structure",
                    "Condition of supports",
                    "Overall bridge condition score",
                    "Vehicles per day")
  )
)
```

These variables help show the conditions of each bridge. Lower ratings and lower sufficiency scores mean that there is a greater chance for structural problems. Bridge age is also an important factor because older bridges are more likely to show wear over time.

There is a possible issue with the dataset missing data. Some bridges might not have complete ratings for every condition measure. If they have missing data, those bridges will be excluded form the analysis or reviewed carefully so that the results aren't misleading.

---

# Source 2: Nebraska Department of Transportation {.unnumbered}

These state bridge reports are from NDOT [@ndot_bridge]. This source adds more local information about repairs and maintenance plans, which can help when looking at short term risk. The main variables used fomr this source are shown in @tbl-ndot.

## Important Variables {.unnumbered}

```{r}
#| echo: false
#| label: tbl-ndot
#| tbl-cap: "Variables used from Nebraska Department of Transportation bridge data."

kable(
  data.frame(
    Variable = c("Planned Repairs", "Funding Levels"),
    Data_Type = c("Categorical (Yes/No)", "Numeric"),
    Description = c("Shows whether repairs are scheduled",
                    "Shows maintenance investment")
  )
)
```

This source is useful because a bridge with planned repairs or available funding may have lower short-term risk than a bridge with no repair plans. It also helps add more local context that may not be fully shown in the national bridge inventory data.

---

# Source 3: Traffic Data {.unnumbered}

This is traffic data that comes from FHWA Highway Statistics. [@fhwa_stats]. This source is mainly used to help estimate the consequences of a bridge collapse instead of the probability of one actually happening.The main variables used from this source are shown in @tbl-traffic.

## Important Variables {.unnumbered}

```{r}
#| echo: false
#| label: tbl-traffic
#| tbl-cap: "Variables used from FHWA traffic data."

kable(
  data.frame(
    Variable = c("Average Annual Daily Traffic (AADT)", "Truck Percentage"),
    Data_Type = c("Numeric", "Numeric (%)"),
    Description = c("Vehicles per day",
                    "Percentage of trucks in traffic")
  )
)
```

This data helps show how serious the consequences would be if a bridge collapsed. A bridge that carries a high amount of traffic, especially truck traffic, would more likely cause more problems.

---

# Is this enough data? {.unnumbered}

I do think it is enough data. Each source provides me with different things that could help assess and prevent the bridge collapse risk in Omaha over the next five years. The NBI data provides the probability of a bridge collapsing. The NDOT data helps with local repair information, and the traffic data measure the impact. I think they work together very well.

---
