# Methods

This analysis uses data from the National Bridge Inventory (NBI), filtered to include only bridges located in Douglas County, Nebraska, which also represents the Omaha area. Bridges with missing structural condition ratings were excluded from the analysis to avoid bias due to incomplete observations.

The main variables used include bridge age, deck rating, superstructure rating, substructure rating, lowest condition rating, and average daily traffic (ADT). Bridge age was calculated using the year built.

This analysis is descriptive and does not directly estimate the probability of bridge failure. Instead, bridge condition ratings are used as a proxy for relative failure risk. Bridges with lower condition ratings are treated as having higher risk, but this does not actually imply a specific probability of collapse.

To evaluate potential consequences of failure, average daily traffic was used as a measure of impact. Bridges with higher traffic volumes were considered to have greater potential consequences if a failure were to happen.

The overall goal of this approach is to identify bridges that may require additional monitoring or maintenance by combining information about structural condition and traffic levels.

### National Bridge Inventory

The data comes from the Federal Highway Administration's National Bridge Inventory [@fhwa_nbi]. This data set provides information about bridge age, condition ratings, location, and traffic levels. The main variables used from this data set are shown in @tbl-nbi.

```{r}
#| echo: false
#| label: tbl-nbi
#| tbl-cap: "Variables used from the National Bridge Inventory dataset."

library(knitr)

kable(
  data.frame(
    Variable = c("Year Built", "Deck Rating", "Superstructure Rating",
                 "Substructure Rating", "Lowest Rating", "Bridge Condition",
                 "Average Daily Traffic"),
    Data_Type = c("Numeric", "Ordinal (0–9)", "Ordinal (0–9)",
                  "Ordinal (0–9)", "Ordinal (0–9)", "Categorical",
                  "Numeric"),
    Description = c("Used to calculate bridge age",
                    "Condition of the bridge deck",
                    "Condition of load-bearing structure",
                    "Condition of bridge supports",
                    "Lowest condition rating among major bridge parts",
                    "Overall condition category such as Good, Fair, or Poor",
                    "Vehicles per day")
  )
)
```

The variables shown in @tbl-nbi provide information on bridge condition, age, and traffic levels. Lower condition ratings indicate a greater likelihood of structural deterioration. Bridge age is also an important factor because older bridges are more likely to show wear over time.

There is a possible issue with missing data in the data set. Some bridges may not have complete ratings for every condition measure. If data is missing, those bridges will either be removed from the analysis or reviewed carefully so the results are not misleading.

### Nebraska Department of Transportation

These state bridge reports are from NDOT [@ndot_bridge]. This source adds more local information about repairs and maintenance plans, which can help when looking at short term risk. The main variables used from this source are shown in @tbl-ndot.

```{r}
#| echo: false
#| label: tbl-ndot
#| tbl-cap: "Variables used from Nebraska Department of Transportation bridge data."

kable(
  data.frame(
    Variable = c("Planned Repairs", "Funding Levels"),
    Data_Type = c("Categorical (Yes/No)", "Numeric"),
    Description = c("Shows whether repairs are scheduled",
                    "Shows maintenance investment")
  )
)
```

The variables in @tbl-ndot provide information on planned repairs and funding, which help interpret short-term maintenance needs.

### Traffic Data

This is traffic data that comes from FHWA Highway Statistics [@fhwa_stats]. This source is mainly used to help estimate the consequences of a bridge collapse instead of the likelihood of failure. The main variables used from this source are shown in @tbl-traffic.

```{r}
#| echo: false
#| label: tbl-traffic
#| tbl-cap: "Variables used from FHWA traffic data."

kable(
  data.frame(
    Variable = c("Average Annual Daily Traffic (AADT)", "Truck Percentage"),
    Data_Type = c("Numeric", "Numeric (%)"),
    Description = c("Vehicles per day",
                    "Percentage of trucks in traffic")
  )
)
```

The data in @tbl-traffic provides a measure of the potential consequences of bridge failure. A bridge that carries a high volume of traffic, especially truck traffic, would more likely cause more problems.

## Selection of Relevant Variables

The variables used in this analysis are shown in @tbl-nbi, @tbl-ndot, and @tbl-traffic.
The main variables include bridge age, deck rating, superstructure rating, substructure rating, lowest condition rating, bridge condition, and average daily traffic.

Bridge age was calculated using the year built.

Bridge condition was measured using deck, superstructure, and substructure ratings.

A bridge was classified as being in poorer condition if its lowest condition rating was 4 or below.

## Evaluating Risk


To evaluate relative failure risk, bridges with lower condition ratings were treated as higher risk. To estimate the consequences of failure, average daily traffic was used as the measure of impact, with higher traffic indicating greater potential disruption.

Overall bridge risk was estimated using the formula:

**Risk = Relative Failure Risk × Consequences of Failure**

This approach was used to identify bridges that may require additional monitoring or maintenance, as discussed in @sec-results.
