```{r}
#| echo: false
#| message: false
#| warning: false
library(dplyr)

omaha_raw <- read.csv("omaha_sample.csv")

omaha_bridges_clean <- omaha_raw |>
  mutate(
    bridge_age = 2026 - year_built,
    lowest_rating = as.numeric(lowest_rating),
    daily_traffic = as.numeric(daily_traffic),
    poor_condition = lowest_rating <= 4
  )
```

# Results {#sec-results}

## Data

```{r}
#| echo: false
#| label: tbl-summary
#| tbl-cap: "Summary statistics for Omaha bridges."

library(dplyr)
library(knitr)

summary_table <- data.frame(
  Statistic = c(
    "Number of bridges",
    "Average bridge age",
    "Poor-condition bridges",
    "Percent poor condition",
    "Maximum daily traffic"
  ),
  Value = c(
    nrow(omaha_bridges_clean),
    round(mean(omaha_bridges_clean$bridge_age, na.rm = TRUE), 1),
    sum(omaha_bridges_clean$poor_condition, na.rm = TRUE),
    paste0(round(mean(omaha_bridges_clean$poor_condition, na.rm = TRUE) * 100, 1), "%"),
    max(omaha_bridges_clean$daily_traffic, na.rm = TRUE)
  )
)

kable(summary_table)
```
The summary statistics in @tbl-summary provide an overview of the Omaha bridge data set. A total of 511 bridges were included in the analysis, with an average age of approximately 45 years. Only 16 bridges, or about 3%, were classified as being in poor condition. These statistics indicate that most bridges are in acceptable condition, although a small subset may require additional attention.

```{r}
#| echo: false
#| label: fig-age
#| fig-cap: "Distribution of bridge age in Omaha."

hist(omaha_bridges_clean$bridge_age,
     xlab = "Bridge Age (years)",
     ylab = "Number of Bridges",
     main = "Bridge Age Distribution")
```

The distribution of bridge age is shown in @fig-age. The data set includes both newer and older bridges, which supports using bridge age as part of the analysis.

```{r}
#| echo: false
#| label: fig-condition
#| fig-cap: "Distribution of bridge condition ratings."

hist(omaha_bridges_clean$lowest_rating,
     xlab = "Condition Rating",
     ylab = "Number of Bridges",
     main = "Bridge Condition Ratings")
```

The distribution of condition ratings is shown in @fig-condition, with most bridges falling into fair to good categories.

Out of the 511 bridges analyzed, 16 bridges have a condition rating of 4 or below, indicating poorer condition. This represents about 3% of all bridges, suggesting that most bridges are not currently in poor condition.

Traffic levels vary widely across bridges, with some carrying only a small number of vehicles per day and others exceeding 100,000 vehicles per day.

## Relative Failure Risk


Bridges with lower condition ratings are identified as having higher relative failure risk based on the classification used in this analysis. Bridges with a lowest condition rating of 4 or below were identified as higher risk bridges.

Older bridges tend to have lower condition ratings, suggesting potential deterioration over time due to wear and usage. While most bridges are still in acceptable condition, there is a small group of bridges with lower ratings that may be at higher risk.

## Consequences of Failure

```{r}
#| echo: false
#| label: fig-risk-scatter
#| fig-cap: "Relationship between bridge condition and traffic volume."

plot(
  omaha_bridges_clean$lowest_rating,
  omaha_bridges_clean$daily_traffic,
  xlab = "Condition Rating",
  ylab = "Daily Traffic",
  main = "Condition vs Traffic"
)
```

The relationship between bridge condition and traffic volume is shown in @fig-risk-scatter. Bridges with both lower condition ratings and higher traffic volumes represent the highest overall risk, as they combine structural concern with greater potential impact.

Bridges with higher daily traffic volumes would result in the largest disruptions if a failure were to occur. Bridges that carry very high traffic volumes, especially on major roads like I-80, would have the greatest impact on transportation and public safety.

Information from NDOT [@ndot_bridge] provides context on maintenance and planned repairs, which may help reduce short-term risk.

These higher risk bridges are typically located along major transportation routes in Omaha. Since these routes are heavily used, a failure in these areas would cause major traffic disruptions and safety concerns.

These findings reinforce the importance of prioritizing bridges that combine both structural concerns and high usage.

### Calculation of Bridge Risk

Overall bridge risk combines relative failure risk and the consequences of failure:

**Risk = Relative Failure Risk × Consequence of Failure**

As shown in @fig-risk, bridge condition is used as a proxy for relative failure risk, and traffic data is used to estimate the impact of a failure.

::: {#fig-risk}

```{mermaid}
flowchart TD
A["Condition"] --> B["Failure Risk"]
C["Traffic"] --> D["Impact"]
B --> E["Overall Risk"]
D --> E
```

Shows how bridge risk is calculated using condition and traffic data.

:::


#### Key Findings

- Most bridges in Omaha are in **fair to good condition**
- Only **16 out of 511 bridges (~3%)** are in poorer condition and may need monitoring or repair
- **Older bridges tend to have lower condition ratings**, showing gradual deterioration
- Bridges with **high traffic volumes** would cause the biggest disruptions if they failed
- Bridges with **both low condition ratings and high traffic** represent the highest risk
